Clinical prediction using similarity-based cohorts: A scoping review

### 2023-03-17 docs/Manuscript/Manuscript.Rmd
add cohen details

---
title: "Clinical prediction using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: 
    affiliation: Some Institute of Technology
    footnote: 1
  - name: Patti McCall
    email: 
    affiliation: Some Institute of Technology
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: Laboratory for Systems Medicine, University of Florida
    correspondingauthor: true
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Laboratory for Systems Medicine, University of Florida
    address: P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
footnote:
  - code: 1
    text: "This is an illustrative author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - case-based reasoning
  - localized modeling
  - nearest neighbors
journal: "Artificial Intelligence In Medicine"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: mybibfile.bib, SR-bibliography,
  SR-0-Found, SR-2-Included, SR-3-Tracked
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    latex_engine: xelatex
    citation_package: natbib
---

@@

One memorable aspiration of clinical informatics was the “green button” that would, in response to clinical queries, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient @Longhurst2014. Localized modeling has recently been put forth as one answer to this appeal. Studies using localized models have been limited by computational cost—a naïve implementation involves fitting, optimizing, and evaluating a model on a separate population for each training case—and recent studies have described algorithms and implementations designed to reduce this cost. As greater access to health data, informatical knowledge, algorithmic efficiency, and computational power continue to spur innovation in CDS, we may expect localized modeling to become more accessible and efficient, its advantages better understood, and its disadvantages ameliorated.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in the development of expert systems [@Aamodt1994]. Early implementations realized a general workflow described as “the four REs”, later the “R4 cycle” [@Aamodt1994; @Begum2011]: Given a new case (or problem), the system retrieves one or more past cases (solved problems) from a corpus, reuses these to generate an understanding of (solution to) the new case, revises this understanding to better fit the new case (sometimes called “adaptation”), and retains the new case and its eventual understanding in the corpus to be retrieved and reused in future. In contrast to rule-based systems, which are variable-based and generally interpretable as a single rule applied to all new cases, case-based systems provide case-specific interpretations in the form of a number of more fully understood reference cases. While the similarity measure used in the retrieval step need not be changed as the corpus grows, proposed measures have been diverse, contested, and rarely systematically compared.

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and Mathematical Reviews. We finalized the PubMed search first, then adapted it to the other platforms. Through each platform, we searched those databases included by default.[^databases]

@@

The structure and terms of our search strings were inspired in part by previous reviews adjacent to our topic of interest [@Sharafoddini2017; @Parimbelli2018]. We organized the PubMed search in disjunctive normal form (an OR of ANDs). Following the solidification of an outline, we expanded each term to include synonyms that are similar enough to be applicable to our search. We then evaluated the expanded search string using the PubMed Advanced Search platform. We initially included each term and their synonyms separately to evaluate what resulted. We pruned several terms that returned no results (“phrases not found”) or to reduce the number of results. At the conclusions of this process for each individual term, we combined the search terms using disjunctive normal form.

@@

Another dominant characteristic of included studies is their remit. We categorized studies as “science” or “quality improvement” based on whether their aim was to produce generalizable knowledge or to improve practice. We classified 7 of the 25 studies as “science”, making the majority of studies from a “quality improvement” standpoint. These studies aimed primarily to provide providers with tools to use in clinic or to improve outcomes for patients currently receiving care. Lastly, we observed a commonality among 6 of the 25 included studies in having evaluated methods using leave-one-out cross-validation (LOOCV). The purpose of this measure is to estimate the overall performance of certain factors when used to make predictions, particularly utilized on smaller data sets, where models benefit greatly from larger training sets and additional model fitting is less costly.

@@

While several studies emphasized the potential or actual value to decision-making of their methods and tools, only one incorporated treatment decisions into their approach: By taking “decision points” as their units of analysis, @Tang2021d and @Ng2021b built not classifiers or predictors but comparative effectiveness models into a localized framework, providing for the first time in our sample explicitly prescriptive rather than descriptive clinical decision support.

@@

CBR can be understood as the opposite side of a trade-off with rule-based reasoning between model size and model complexity; @Doyle2006 describe their approach as “knowledge-light”, in that “the cases do not contain explicit explanation structures; instead, explanation is achieved by comparison of the query case with retrieved cases”. This means that the greatest performance and efficiency challenges in CBR have to do with the retrieval and revision phases in the R4 cycle. Several studies addressed these challenges: @Park2006, while excluded from the synthesis, were the earliest to propose that cohorts be bounded by a similarity threshold rather than by a number of cases, and this improved performance in their experiments. @CampilloGimenez2013 leveraged predictor weights obtained from logistic regression to inform the similarity calculation used in retrieval. @Lowsky2013 used non-parametric models to reduce the computational burden of revision (prediction). @Ma2020a proposed to improve efficiency along the entire R4 cycle, but in particular by only executing task-dependent steps in real time (“just-in-time learning”), and @Ng2021b and @Tang2021d partitioned multiple phases in the R4 cycle into offline and online components, only the latter of which would be performed in real time as new data are received.

@@

These programs used significantly varying terminology for common concepts, and in particular no common term is in use for what we here refer to as localized modeling; authors described this kind of method as “targeted prognosis” [@Mariuzzi1997], “transductive inference” [@Song2006c], “personalized decision support” [@Lee2015], “personalized (predictive) models” [@Liang2015, in contrast to “local models”; @Ng2015; @Wang2019; @Ma2020a], and “precision cohort” analysis [@Wang2019; @Tang2021d; @Ng2021b]. We find uses of “targeted”, “personalized”, and “precision” generic and imprecise to this approach, while transductive inference is an established term for a broader set of methods in ML. Because local naturally contrasts with global, we propose “localized models” as a suitable term for this counterpart to globally-fitted models.

@@

Distinct from but related to similarity learning was supervised cohort construction. This was an adaptive step taken by @Park2006 to optimize the predictive performance of models fitted to cohorts retrieved using a fixed similarity rather than count threshold, a counterpart to conventional CBR they called “statistical CBR”. The step used a heuristic procedure to locate a similarity threshold that (locally) maximizes predictive accuracy on the training set. This is analogous to optimizing the neighborhood size parameter in NN predictive modeling, so we will not discuss it further except to mention that the authors found statistical CBR to outperform conventional CBR on several data sets. @CampilloGimenez2012a combined these learning techniques: They used logistic regression on the training set to obtain weights for the predictors (by predicting the binary outcome of kidney transplant waitlist registration) and for the cases (by predicting agreement of outcome between an index case and other training cases), and they used exhaustion to optimize the size of the retrieved cohort for the accuracy of the NN model using the previously optimized weights.

@@

We propose the term _localized modeling_ to encompass an approach derived from CBR in which parameterized models are fitted in a standardized way to nearest neighborhoods of past or training cases according to a measure of patient similarity. We conducted a systematic search for studies that apply localized models to tasks involving health data and synthesized these largely independently developed approaches into a general framework. While the search was limited by low inter-rater reliability and failure to recover several motivating examples, the included studies used many of the same underlying/fundamental tools to build, optimize, and evaluate their methods. We therefore believe that our framework can serve to taxonomize ongoing work of this type and inform the development of flexible/customizable implementations. Indeed, the availability of increasing computational power, the diversity of tasks to which these models were applied and of technical specifications they employed, and the apparent lack of any multi-group research program to date suggest great potential for growth. Whereas precious few of the reviewed studies used these models, for any task other than prediction, despite widespread suspicion among clinicians of “black-box” models and growing interest in interpretable alternatives, we recommend that future work put greater emphasis on the development and validation of interpretable localized models and on their reception by communities of medical research and clinical practice.

>>>>

---
title: "Clinical prediction using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall
    email: 
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: Nova Southeastern University College of Osteopathic Medicine, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Department, Street, City, State, Zip
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
journal: "Artificial Intelligence In Medicine"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: SR-bibliography,
  SR-0-Found, SR-2-Included, SR-3-Tracked
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    latex_engine: xelatex
    citation_package: natbib
---

@@

One memorable aspiration of clinical informatics was the "green button" that would, in response to clinical queries, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient @Longhurst2014. Localized modeling has recently been put forth as one answer to this appeal. Studies using localized models have been limited by computational cost—a naïve implementation involves fitting, optimizing, and evaluating a model on a separate population for each training case—and recent studies have described algorithms and implementations designed to reduce this cost. As greater access to health data, informatical knowledge, algorithmic efficiency, and computational power continue to spur innovation in CDS, we may expect localized modeling to become more accessible and efficient, its advantages better understood, and its disadvantages ameliorated.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in the development of expert systems [@Aamodt1994]. Early implementations realized a general workflow described as "the four REs", later the "R4 cycle" [@Aamodt1994; @Begum2011]: Given a new case (or problem), the system retrieves one or more past cases (solved problems) from a corpus, reuses these to generate an understanding of (solution to) the new case, revises this understanding to better fit the new case (sometimes called "adaptation"), and retains the new case and its eventual understanding in the corpus to be retrieved and reused in future. In contrast to rule-based systems, which are variable-based and generally interpretable as a single rule applied to all new cases, case-based systems provide case-specific interpretations in the form of a number of more fully understood reference cases. While the similarity measure used in the retrieval step need not be changed as the corpus grows, proposed measures have been diverse, contested, and rarely systematically compared.

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and MathSciNet. We finalized the PubMed search first, then adapted it to the other platforms. Through each platform, we searched those databases included by default.[^databases]

@@

The structure and terms of our search strings were inspired in part by previous reviews adjacent to our topic of interest [@Sharafoddini2017; @Parimbelli2018]. We organized the PubMed search in disjunctive normal form (an OR of ANDs). Following the solidification of an outline, we expanded each term to include synonyms that are similar enough to be applicable to our search. We then evaluated the expanded search string using the PubMed Advanced Search platform. We initially included each term and their synonyms separately to evaluate what resulted. We pruned several terms that returned no results ("phrases not found") or to reduce the number of results. At the conclusions of this process for each individual term, we combined the search terms using disjunctive normal form.

@@

Another dominant characteristic of included studies is their remit. We categorized studies as "science" or "quality improvement" based on whether their aim was to produce generalizable knowledge or to improve practice. We classified 7 of the 25 studies as "science", making the majority of studies from a "quality improvement" standpoint. These studies aimed primarily to provide providers with tools to use in clinic or to improve outcomes for patients currently receiving care. Lastly, we observed a commonality among 6 of the 25 included studies in having evaluated methods using leave-one-out cross-validation (LOOCV). The purpose of this measure is to estimate the overall performance of certain factors when used to make predictions, particularly utilized on smaller data sets, where models benefit greatly from larger training sets and additional model fitting is less costly.

@@

While several studies emphasized the potential or actual value to decision-making of their methods and tools, only one incorporated treatment decisions into their approach: By taking "decision points" as their units of analysis, @Tang2021d and @Ng2021b built not classifiers or predictors but comparative effectiveness models into a localized framework, providing for the first time in our sample explicitly prescriptive rather than descriptive clinical decision support.

@@

CBR can be understood as the opposite side of a trade-off with rule-based reasoning between model size and model complexity; @Doyle2006 describe their approach as "knowledge-light", in that "the cases do not contain explicit explanation structures; instead, explanation is achieved by comparison of the query case with retrieved cases". This means that the greatest performance and efficiency challenges in CBR have to do with the retrieval and revision phases in the R4 cycle. Several studies addressed these challenges: @Park2006, while excluded from the synthesis, were the earliest to propose that cohorts be bounded by a similarity threshold rather than by a number of cases, and this improved performance in their experiments. @CampilloGimenez2013 leveraged predictor weights obtained from logistic regression to inform the similarity calculation used in retrieval. @Lowsky2013 used non-parametric models to reduce the computational burden of revision (prediction). @Ma2020a proposed to improve efficiency along the entire R4 cycle, but in particular by only executing task-dependent steps in real time ("just-in-time learning"), and @Ng2021b and @Tang2021d partitioned multiple phases in the R4 cycle into offline and online components, only the latter of which would be performed in real time as new data are received.

@@

These programs used significantly varying terminology for common concepts, and in particular no common term is in use for what we here refer to as localized modeling; authors described this kind of method as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006c], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020a], and "precision cohort" analysis [@Wang2019; @Tang2021d; @Ng2021b]. We find uses of "targeted", "personalized", and "precision" generic and imprecise to this approach, while transductive inference is an established term for a broader set of methods in ML. Because local naturally contrasts with global, we propose "localized models" as a suitable term for this counterpart to globally-fitted models.

@@

Distinct from but related to similarity learning was supervised cohort construction. This was an adaptive step taken by @Park2006 to optimize the predictive performance of models fitted to cohorts retrieved using a fixed similarity rather than count threshold, a counterpart to conventional CBR they called "statistical CBR". The step used a heuristic procedure to locate a similarity threshold that (locally) maximizes predictive accuracy on the training set. This is analogous to optimizing the neighborhood size parameter in NN predictive modeling, so we will not discuss it further except to mention that the authors found statistical CBR to outperform conventional CBR on several data sets. @CampilloGimenez2012a combined these learning techniques: They used logistic regression on the training set to obtain weights for the predictors (by predicting the binary outcome of kidney transplant waitlist registration) and for the cases (by predicting agreement of outcome between an index case and other training cases), and they used exhaustion to optimize the size of the retrieved cohort for the accuracy of the NN model using the previously optimized weights.

@@

We propose the term _localized modeling_ to encompass an approach derived from CBR in which parameterized models are fitted in a standardized way to nearest neighborhoods of past or training cases according to a measure of patient similarity. We conducted a systematic search for studies that apply localized models to tasks involving health data and synthesized these largely independently developed approaches into a general framework. While the search was limited by low inter-rater reliability and failure to recover several motivating examples, the included studies used many of the same underlying/fundamental tools to build, optimize, and evaluate their methods. We therefore believe that our framework can serve to taxonomize ongoing work of this type and inform the development of flexible/customizable implementations. Indeed, the availability of increasing computational power, the diversity of tasks to which these models were applied and of technical specifications they employed, and the apparent lack of any multi-group research program to date suggest great potential for growth. Whereas precious few of the reviewed studies used these models, for any task other than prediction, despite widespread suspicion among clinicians of "black-box" models and growing interest in interpretable alternatives, we recommend that future work put greater emphasis on the development and validation of interpretable localized models and on their reception by communities of medical research and clinical practice.

## Appendix

### Search strings

Here we reproduce the search strings and platform specifications used in our literature search.
We first finalized the **PubMed** search string below:

```
(
  "case-based reasoning" [All Fields] OR
  "case-based system" [All Fields]
) OR (
  "individualized modeling" [All Fields] OR
  "personalized modeling" [All Fields] OR
  "customized modeling" [All Fields]
) OR (
  "individualized cohort" [All Fields]
) OR (
  (
    "patient similarity" [All Fields] OR
    "patient distance" [All Fields] OR
    "patient connection" [All Fields] OR
    "patient affinity" [All Fields] OR
    "patient clustering" [All Fields]
  ) AND (
    "cohort study" [All Fields]
  )
)
```

This search yielded 423 results.

We then generated analogous search strings or search strategies for the Web of Science, Academic Search Premier, and MathSciNet platforms, based on the logics and syntaxes of their respective interfaces.

For **Web of Science**, we searched for several separate disjunctions derived from the PubMed search string.
The separate search strings and the number of results obtained using each are below.

------------------------------------------------
Search string                    Number of items
------------------------------- ----------------
`"case-based reasoning" OR`                     
`"case-based system"`                      3,636

`"individualized model" OR`                  444
`"individualized modeling" OR`                  
`"personalized model" OR`                    
`"personalized modeling" OR`                    
`"customized model"`                         
`"customized modeling"`

`"individualized cohort"`                      1

`"patient similarity" OR`                     48
`"patient distance" OR`                         
`"patient connection" OR`                       
`"patient affinity" OR`                         
`"patient clustering"`                          
Refined search: `"cohort"`                      
------------------------------------------------

We found the results of the first search string to be predominantly irrelevant. To reduce review time, these were dropped.
The last search was refined with an additional term following the initial disjunctive search.
Our searches on Web of Science thus yielded 493 results.

When searching **Academic Search Premier**, we checked the option "Scholarly (Peer-Reviewed) Journals", unchecked the option "Apply equivalent subjects", and searched for several separate disjunctions and conjunctions of strings in the "TX All Text" field.
We obtained the resulting citations via email in RIS format.

------------------------------------------------
Search string                    Number of items
------------------------------- ----------------
`"case-based reasoning" OR`                     
`"case-based system"`                      3,283

`"individualized modeling" OR`               100
`"personalized modeling" OR`                    
`"customized modeling"`

`"individualized cohort"`                      2

`"patient similarity" AND`                    14
`"cohort study"`

`"patient distance" AND`                      15
`"cohort study"`

`"patient connection" AND`                     6
`"cohort study"`

`"patient affinity" AND`                       0
`"cohort study"`

`"patient clustering" AND`                    50
`"cohort study"`                                
------------------------------------------------

We found the results of the first search string to be predominantly irrelevant. To reduce review time, these were dropped.
Our searches on Academic Search Premier thus yielded 187 results.

For **MathSciNet**, the portal to the _Mathematical Reviews_ database, we used the same separate searches as for Web of Science, in some cases expanded to obtain more results. Those which yielded nonzero numbers of results are below:

------------------------------------------------
Search string                    Number of items
------------------------------- ----------------
`"case-based reasoning"`                     110

`"individualized model`                        2

`"patient similarity" AND`                     1
`"cohort"`                                
------------------------------------------------

Our searches of _Mathematical Reviews_ therefore yielded 113 results.

These totaled 1,422 sources from all platforms. We organized the full results in a Zotero collection alongside the seed set and created a single folder for the 25 results reviewed in detail.
